% ====================================================================
\section{Algorithms and Implementation}
\label{sec:algorithm}
% ====================================================================

The estimation procedure consists of two stages.
Algorithm~\ref{alg:ilr_egsca} computes the globally optimal spectral
solution for the profiled linearised objective; its output is
deterministic and non-iterative.
Algorithm~\ref{alg:refine} is an optional single-pass refinement that
re-estimates the topic--term distributions outside the ILR subspace,
improving topic exclusivity without modifying the document--topic
proportions or path coefficients.

% ---- Algorithm 1 --------------------------------------------------
\begin{algorithm}[h]
\caption{ILR-EGSCA Spectral Fit}
\label{alg:ilr_egsca}
\begin{algorithmic}[1]
\REQUIRE Document--term matrix
  $\mathbf{W}\in\mathbb{R}^{M\times N}_{\geq 0}$,
  column-centred covariate matrix
  $\mathbf{C}\in\mathbb{R}^{M\times P}$,
  number of topics $K$,
  regularisation $\lambda>0$,
  ILR contrast matrix
  $\mathbf{V}\in\mathbb{R}^{K\times(K-1)}$.
\ENSURE Topic proportions $\hat{\boldsymbol{\Pi}}\in\mathbb{R}^{M\times K}$,
  ILR loadings $\hat{\boldsymbol{\Psi}}\in\mathbb{R}^{(K-1)\times N}$,
  ILR path coefficients $\hat{\mathbf{B}}_z\in\mathbb{R}^{P\times(K-1)}$.

\medskip
\STATE \textbf{Centre.}\;
  $\widetilde{\mathbf{W}} \leftarrow \mathbf{W}
    - M^{-1}\mathbf{1}_M\mathbf{1}_M^{\!\top}\mathbf{W}$
  \hfill$\triangleright$ Prop.~\ref{prop:mu_profile}

\STATE \textbf{Truncated SVD.}\;
  $\mathbf{U}\boldsymbol{\Sigma}\mathbf{V}_W^{\!\top}
    \leftarrow \text{SVD}_r(\widetilde{\mathbf{W}})$
  \hfill$\triangleright$ randomised, rank $r$

\STATE \textbf{Covariate QR.}\;
  $\mathbf{Q}_C\mathbf{R}_C \leftarrow \text{QR}(\mathbf{C})$

\STATE \textbf{Augmented factor.}\;
  $\mathbf{H} \leftarrow [\,\mathbf{U}\boldsymbol{\Sigma}
    \;\;\;\sqrt{\lambda}\,\mathbf{Q}_C\,]
    \in\mathbb{R}^{M\times(r+P)}$
  \hfill$\triangleright$ Eq.~\eqref{eq:H_matrix}

\STATE \textbf{Reduced eigendecomposition.}\;
  $\mathbf{H}^{\!\top}\mathbf{H}
    = \mathbf{E}\,\diag(\sigma_1,\dots,\sigma_{r+P})\,
      \mathbf{E}^{\!\top}$
  \hfill$\triangleright$ $(r{+}P)\times(r{+}P)$

\STATE \textbf{ILR topic scores.}\;
  $\mathbf{Z}^* \leftarrow \mathbf{H}\mathbf{E}_{1:K-1}\,
    \diag(\sigma_1^{-1/2},\dots,\sigma_{K-1}^{-1/2})$
  \hfill$\triangleright$ Thm.~\ref{thm:spectral}

\STATE \textbf{ILR loadings.}\;
  $\hat{\boldsymbol{\Psi}} \leftarrow
    K\,\mathbf{Z}^{*\top}\widetilde{\mathbf{W}}$
  \hfill$\triangleright$ Prop.~\ref{prop:Psi_profile}

\STATE \textbf{Varimax rotation} (optional)\textbf{.}\;
  $\mathbf{R}^* \leftarrow \mathrm{varimax}
    (\hat{\boldsymbol{\Psi}}^{\!\top})$;\;
  $\mathbf{Z}^* \leftarrow \mathbf{Z}^*\mathbf{R}^*$;\;
  $\hat{\boldsymbol{\Psi}} \leftarrow
    \mathbf{R}^{*\top}\hat{\boldsymbol{\Psi}}$
  \hfill$\triangleright$ Prop.~\ref{prop:rotation_invariance}

\STATE \textbf{Path coefficients.}\;
  $\hat{\mathbf{B}}_z \leftarrow
    (\mathbf{C}^{\!\top}\mathbf{C})^{-1}
    \mathbf{C}^{\!\top}\mathbf{Z}^*$
  \hfill$\triangleright$ Prop.~\ref{prop:Bz_profile}

\STATE \textbf{Topic proportions.}\;
  $\hat{\pi}_{ik} \leftarrow
    \exp\!\bigl((\mathbf{V}\mathbf{z}_i^*)_k\bigr)\big/
    \sum_{k'}\exp\!\bigl((\mathbf{V}\mathbf{z}_i^*)_{k'}\bigr)$
  \quad$\forall\; i,k$
  \hfill$\triangleright$ ILR inverse

\RETURN $\hat{\boldsymbol{\Pi}},\;
  \hat{\boldsymbol{\Psi}},\;
  \hat{\mathbf{B}}_z,\;
  \mathbf{Z}^*$
\end{algorithmic}
\end{algorithm}

\noindent
Algorithm~\ref{alg:ilr_egsca} is \emph{non-iterative}: it consists
of a fixed sequence of matrix decompositions and linear-algebra
operations, with no convergence loop and no random initialisation.
Its output is deterministic for any given input (up to the optional
varimax step, which is itself a deterministic function of
$\hat{\boldsymbol{\Psi}}$). The computational bottleneck is the
truncated SVD in Step~2, implemented via randomised projection
\citep{halko2011finding} at cost $O(MNr)$; all subsequent steps
operate on matrices of dimension at most
$(r{+}P)\times(r{+}P)$.


% ---- Structural limitation of Phi ---------------------------------
\subsection{Topic--Term Refinement}
\label{subsec:refinement}

The ILR-based topic--term matrix
$\hat{\boldsymbol{\Phi}} = \mathbf{V}\hat{\boldsymbol{\Psi}}
+ \mathbf{1}_K\bar{\mathbf{w}}^{\!\top}$
(Corollary~\ref{cor:recovery}) constrains the $K$~topic--term
distributions to a $(K{-}1)$-dimensional affine subspace of
$\mathbb{R}^N$ anchored at the corpus mean~$\bar{\mathbf{w}}$.
This is the algebraic cost of the closed-form solution: the same
structure that permits profiling out $\boldsymbol{\Psi}$ in
Proposition~\ref{prop:Psi_profile} forces the $K$~rows of
$\hat{\boldsymbol{\Phi}}$ to share a common baseline and
differentiate only along $K{-}1$ orthogonal directions. The result is
high topic similarity and limited exclusivity relative to methods
that estimate each topic distribution freely (e.g., STM, LDA).

To address this, we introduce a single-pass refinement that
re-estimates $\hat{\boldsymbol{\Phi}}$ outside the ILR subspace
while leaving the spectral estimates
$(\hat{\boldsymbol{\Pi}}, \hat{\mathbf{B}}_z, \mathbf{Z}^*)$
unchanged. The key observation is that the document--topic
proportions $\hat{\boldsymbol{\Pi}}$ produced by
Algorithm~\ref{alg:ilr_egsca}---which lie on the simplex and enjoy
the theoretical guarantees of
Section~\ref{sec:theory}---provide a valid soft assignment of
documents to topics, from which unconstrained topic--term
distributions can be recovered in closed form.

\paragraph{Assignment step.}
Two strategies are available for computing the document--topic
assignment weights $\mathbf{Q}\in\mathbb{R}^{M\times K}$:

\begin{enumerate}
  \item[\emph{(a)}]
    \textbf{$K$-means on ILR scores} (default).
    Apply $K$-means clustering to the scaled ILR scores
    $\sqrt{M}\,\mathbf{Z}^*$ (which have approximately unit
    column variance), producing hard assignments
    $q_{ik}\in\{0,1\}$.
    This directly exploits the spectral geometry:
    clusters in ILR space correspond to groups of documents with
    similar topic compositions.

  \item[\emph{(b)}]
    \textbf{Soft E-step}.
    Compute soft assignments via
    $q_{ik}\propto\exp\bigl(\sum_n\widetilde{w}_{in}\,
    \log\hat\phi^{(0)}_{kn}\bigr)$,
    where
    $\hat{\boldsymbol{\Phi}}^{(0)}
    = \softmax(\mathbf{V}\hat{\boldsymbol{\Psi}})$
    is the ILR-based topic--word distribution.
    This is one LDA-style E-step and tends to improve semantic
    coherence at the cost of smaller gains in diversity.
\end{enumerate}

\paragraph{M-step.}
Given $\mathbf{Q}$, the refined topic--term distributions are
\begin{equation}
  \hat{\phi}^{(1)}_{kn}
  \;=\;
  \frac{\sum_{i=1}^M q_{ik}\,\widetilde{w}_{in} + \varepsilon}
       {\sum_{n'=1}^{N}\bigl(\sum_{i=1}^M q_{ik}\,
        \widetilde{w}_{in'} + \varepsilon\bigr)},
  \label{eq:mstep_phi}
\end{equation}
where $\varepsilon>0$ is a Laplace smoothing constant.
In matrix form:
$\hat{\boldsymbol{\Phi}}^{(1)}
 = \mathrm{row\text{-}normalise}\!\bigl(
     \mathbf{Q}^{\!\top}\widetilde{\mathbf{W}} + \varepsilon\bigr)$,
which costs $O(MKN)$---a single matrix multiplication.

\paragraph{Temperature sharpening.}
An optional temperature parameter $\tau\in(0,1]$ further concentrates
mass on the highest-probability terms:
\begin{equation}
  \hat{\phi}^{(\tau)}_{kn}
  \;=\;
  \frac{\bigl(\hat{\phi}^{(1)}_{kn}\bigr)^{1/\tau}}
       {\sum_{n'}\bigl(\hat{\phi}^{(1)}_{kn'}\bigr)^{1/\tau}}.
  \label{eq:temperature}
\end{equation}
Setting $\tau=1$ recovers the M-step distribution unchanged;
$\tau<1$ sharpens the distribution towards a sparse representation.

% ---- Algorithm 2 --------------------------------------------------
\begin{algorithm}[t]
\caption{Topic--Term Refinement (post-spectral)}
\label{alg:refine}
\begin{algorithmic}[1]
\REQUIRE Spectral fit from Algorithm~\ref{alg:ilr_egsca}:
  $\hat{\boldsymbol{\Pi}}$, $\hat{\boldsymbol{\Psi}}$,
  $\mathbf{Z}^*$;\;
  document--term matrix~$\mathbf{W}$;\;
  smoothing $\varepsilon>0$;\;
  temperature $\tau\in(0,1]$;\;
  method $\in\{\text{kmeans},\,\text{em}\}$.
\ENSURE Refined topic--term matrix
  $\hat{\boldsymbol{\Phi}}^{(\tau)}
   \in\mathbb{R}^{K\times N}$.

\medskip
\IF{method $=$ kmeans}
  \STATE $\mathbf{Q} \leftarrow \text{$K$-means}
    (\sqrt{M}\,\mathbf{Z}^*,\; K\text{ clusters})$
  \hfill$\triangleright$ hard assignment
\ELSE
  \STATE $\hat{\boldsymbol{\Phi}}^{(0)} \leftarrow
    \softmax(\mathbf{V}\hat{\boldsymbol{\Psi}})$
  \STATE $q_{ik} \leftarrow
    \softmax_k\!\bigl(\widetilde{\mathbf{w}}_i^{\!\top}
    \log\hat{\boldsymbol{\Phi}}^{(0)\top}\bigr)$
  \hfill$\triangleright$ soft E-step
\ENDIF
\STATE \textbf{M-step.}\;
  $\hat{\boldsymbol{\Phi}}^{(1)} \leftarrow
    \text{row-normalise}\!\bigl(
      \mathbf{Q}^{\!\top}\widetilde{\mathbf{W}}
      + \varepsilon\bigr)$
  \hfill$\triangleright$ Eq.~\eqref{eq:mstep_phi}
\STATE \textbf{Temperature.}\;
  $\hat{\phi}^{(\tau)}_{kn} \leftarrow
    \hat{\phi}^{(1)1/\tau}_{kn}\big/
    \sum_{n'}\hat{\phi}^{(1)1/\tau}_{kn'}$
  \hfill$\triangleright$ Eq.~\eqref{eq:temperature}

\RETURN $\hat{\boldsymbol{\Phi}}^{(\tau)}$
\end{algorithmic}
\end{algorithm}

\begin{remark}[Separation of estimation and refinement]
\label{rem:separation}
Algorithm~\ref{alg:refine} modifies only the topic--term matrix
$\hat{\boldsymbol{\Phi}}$. The document--topic proportions
$\hat{\boldsymbol{\Pi}}$, the path coefficients
$\hat{\mathbf{B}}_z$, and the ILR scores $\mathbf{Z}^*$ are
\emph{unchanged}. Therefore all theoretical guarantees of
Section~\ref{sec:theory}---global optimality
(Theorem~\ref{thm:global}), consistency
(Theorem~\ref{thm:consistency}), asymptotic normality
(Theorem~\ref{thm:normality}), and the linearisation bound
(Proposition~\ref{prop:linearisation_error})---remain valid for the
structural parameters
$(\hat{\boldsymbol{\Pi}},\hat{\mathbf{B}}_z)$ after refinement.
The refinement addresses only the representation of the
topic--term distributions, which are a derived quantity not covered
by the asymptotic theory.
\end{remark}

\begin{remark}[Why a single pass suffices]
\label{rem:single_pass}
In classical EM for topic models, the E-step and M-step are iterated
until convergence. Here, the document--topic proportions are
\emph{fixed} by the spectral solution and are not updated by the
M-step, so there is no E-M loop: the refinement is a single
forward pass from the spectral $\hat{\boldsymbol{\Pi}}$
(or from the $K$-means partition of $\mathbf{Z}^*$)
to the unconstrained $\hat{\boldsymbol{\Phi}}^{(\tau)}$.
The computational cost is dominated by the matrix product
$\mathbf{Q}^{\!\top}\widetilde{\mathbf{W}}\in\mathbb{R}^{K\times N}$
at $O(MKN)$; for $K$-means, the clustering step adds $O(MK^2)$
per restart, which is negligible for moderate~$K$.
\end{remark}


\subsection{Complete Pipeline}
\label{subsec:pipeline}

The recommended workflow for applied use combines both algorithms:
%
\begin{enumerate}
  \item \textbf{Spectral fit}
    (Algorithm~\ref{alg:ilr_egsca}): produces
    $\hat{\boldsymbol{\Pi}}$, $\hat{\mathbf{B}}_z$,
    $\hat{\boldsymbol{\Psi}}$, $\mathbf{Z}^*$ with full theoretical
    guarantees. Varimax rotation is applied by default.

  \item \textbf{Topic--term refinement}
    (Algorithm~\ref{alg:refine}): re-estimates
    $\hat{\boldsymbol{\Phi}}$ via $K$-means on ILR scores followed by
    a weighted M-step with Laplace smoothing and optional temperature
    sharpening. Default: $\varepsilon=10^{-4}$, $\tau=0.5$.

  \item \textbf{Inference}: standard errors for
    $\hat{\mathbf{B}}_z$ are computed from the analytical sandwich
    formula (Theorem~\ref{thm:normality}) or, in the presence of
    small eigengaps, via nonparametric bootstrap.
\end{enumerate}
%
The entire pipeline---from raw document--term matrix to topic
proportions, path coefficients with confidence intervals, and
refined topic--term distributions---requires no iterative
optimisation and no random initialisation beyond the $K$-means
step in Algorithm~\ref{alg:refine}.

\subsection{Computational Complexity}
\label{subsec:complexity}

Table~\ref{tab:complexity} summarises the cost of each step. The
dominant cost is the randomised SVD in Algorithm~\ref{alg:ilr_egsca},
Step~2, which scales as $O(MN\log r)$ via the algorithm
of~\citet{halko2011finding}. All other steps are linear in~$M$.
For comparison, STM requires $O(MK\cdot\text{iter}_{\text{EM}})$
per variational E-step and typically 50--100 EM iterations, making
it one to two orders of magnitude slower on large corpora.

\begin{table}[h]
\centering
\caption{Computational cost per step. Here $r$~is the SVD truncation
rank, $P$~the number of covariates, $K$~the number of topics, and
$N_{\text{start}}$~the number of $K$-means restarts.}
\label{tab:complexity}
\begin{tabular}{llc}
\toprule
Step & Operation & Cost \\
\midrule
\multicolumn{3}{l}{\emph{Algorithm~\ref{alg:ilr_egsca}: Spectral fit}} \\
\; 1 & Column centering & $O(MN)$ \\
\; 2 & Randomised SVD (rank $r$) & $O(MN\log r)$ \\
\; 3 & Covariate QR & $O(MP^2)$ \\
\; 4--6 & Gram eigendecomposition & $O((r{+}P)^3 + M(r{+}P)^2)$ \\
\; 7 & ILR loadings $\hat{\boldsymbol{\Psi}}$ & $O((K{-}1)MN)$ \\
\; 8 & Varimax rotation & $O(N(K{-}1)^2)$ \\
\; 9 & Path coefficients & $O(MP(K{-}1))$ \\
\; 10 & Softmax (ILR inverse) & $O(MK)$ \\
\midrule
\multicolumn{3}{l}{\emph{Algorithm~\ref{alg:refine}: Refinement}} \\
\; 1 & $K$-means on $\mathbf{Z}^*$ & $O(MK^2 N_{\text{start}})$ \\
\; 2 & M-step & $O(MKN)$ \\
\; 3 & Temperature sharpening & $O(KN)$ \\
\bottomrule
\end{tabular}
\end{table}