% ====================================================================
\section{Introduction}
\label{sec:introduction}
% ====================================================================
 
Topic models are a workhorse of modern text analysis. By representing
each document as a mixture of latent themes, they reduce
high-dimensional word-count data to an interpretable low-rank
structure that can be explored, visualised, and related to external
variables. In many applications the central scientific question is not
``what are the topics?'' but ``how does topic prevalence vary with
covariates?''---for instance, how do perceptions of the most important
political issue shift across demographic and attitudinal groups?
 
The Structural Topic Model \citep[STM,][]{roberts2014stm} addresses
this question by embedding a logistic-normal regression of
document--topic proportions on covariates within a hierarchical
generative model, estimated via variational expectation--maximisation.
STM has become the dominant framework for covariate-driven topic
analysis in political science, sociology, and the digital humanities.
Yet its inferential machinery carries three limitations that are
inherent to the variational approach.
First, the optimisation landscape is non-convex: the variational EM
algorithm converges to a local optimum whose quality depends on
initialisation.
Second, the method is iterative: each EM cycle requires a pass over
all documents, and convergence typically demands tens to hundreds of
iterations, making the method slow on large corpora.
Third, uncertainty quantification for the covariate effects requires
parametric bootstrap resampling---a computationally expensive
post-estimation step with no closed-form alternative.
 
This paper shows that all three limitations dissolve when the problem
is reformulated in the natural geometry of the probability simplex.
 
Document--topic proportions live on the $K$-simplex
$\mathcal{S}^K$, a $(K{-}1)$-dimensional manifold whose intrinsic
metric structure is given by the Aitchison inner
product~\citep{aitchison1986statistical}.
The isometric log-ratio (ILR)
transformation~\citep{egozcue2003isometric} maps the simplex to
$\mathbb{R}^{K-1}$ without distortion, replacing the constrained
compositional estimation problem with an unconstrained one in
Euclidean space. We exploit this coordinate change to formulate a
structural topic model whose objective function---combining
document--term reconstruction with a covariate-driven structural
penalty---admits sequential closed-form profiling of all nuisance
parameters, reducing the entire estimation problem to a single
eigenvalue decomposition.
 
The resulting estimator, which we call \emph{egscatm} (exploratory
generalised structured component analysis for topic models), has
several distinctive features.
It is \emph{non-iterative}: the solution is a deterministic sequence
of matrix decompositions with no convergence loop and no random
initialisation.
It is \emph{globally optimal} for the profiled linearised objective:
the Ky~Fan eigenvalue theorem guarantees that no other feasible
parameter set achieves a lower value.
And it yields \emph{analytical standard errors} for the covariate
path coefficients: the asymptotic covariance matrix is derived in
closed form from eigenvector perturbation theory, providing valid
confidence intervals without bootstrap.
 
The key algebraic insight is that profiling out the corpus mean,
the topic--term loadings, and the regression coefficients in
sequence leaves a trace maximisation over an orthonormality-constrained
matrix---exactly the problem solved by the leading eigenvectors of a
covariate-augmented document similarity matrix. This matrix,
$\mathbf{S}_z = \widetilde{\mathbf{W}}\widetilde{\mathbf{W}}^{\!\top}
+ \lambda\mathbf{P}_C$, blends centred document similarity with the
inter-document structure induced by covariates, and the regularisation
parameter $\lambda$ governs the balance: $\lambda=0$ recovers Latent
Semantic Analysis; $\lambda\to\infty$ collapses the topics onto the
covariate space.
 
The spectral solution operates in the ILR coordinate system, where
the topic--term distributions are constrained to a
$(K{-}1)$-dimensional affine subspace anchored at the corpus mean.
This constraint---the algebraic price of the closed-form
solution---limits topic exclusivity relative to methods that
estimate each topic distribution freely.
To address this, we introduce a single-pass post-spectral refinement
that re-estimates the topic--term matrix via a weighted
assignment--maximisation step, breaking the subspace constraint while
leaving the document--topic proportions and path coefficients---and
their theoretical guarantees---intact.
 
We develop the theoretical foundations of the estimator in detail.
Global optimality for the linearised objective is established via the
Ky~Fan theorem. Identifiability conditions under both simple and
degenerate eigenspectra are characterised.
Consistency and first-order asymptotic normality of the path
coefficients are derived via the Davis--Kahan theorem and eigenvector
perturbation expansions, yielding an explicit sandwich covariance
formula.
A linearisation error bound under an eigenvector delocalisation
condition quantifies the approximation quality relative to the
original simplex problem, showing that the per-document discrepancy
vanishes at rate $O(M^{-2})$ as the corpus grows.
 
A three-block simulation study validates the theory: the analytical
confidence intervals achieve nominal coverage without bootstrap; the
linearisation error is three to four orders of magnitude below the
theoretical bound; and the spectral estimator achieves speedups of
$9\times$ to $107\times$ over STM with comparable accuracy in the
recovery of covariate effects under moderate signal strength.
An application to open-ended responses from the British Election
Study demonstrates the method on a substantive political science
question: how do perceptions of the most important issue facing
the country vary across demographic and attitudinal groups?
 
The paper makes three contributions that sit at the intersection
of distinct methodological traditions.
To \emph{topic modeling}, it contributes a non-iterative estimator
with analytical inference---a qualitatively different object from the
variational and sampling-based estimators that dominate the field.
To \emph{structural equation modeling}, it contributes a
compositional latent-variable model where the latent space is not
arbitrary but is the ILR transform of the probability simplex,
giving the component scores a principled interpretation as
document--topic compositions.
To \emph{compositional data analysis}, it contributes a spectral
method for high-dimensional compositions with covariate structure,
extending the Aitchison framework beyond its traditional
low-dimensional settings.
 
The remainder of the paper is organised as follows.
Section~\ref{sec:background} reviews the Aitchison geometry and the
ILR transformation.
Section~\ref{sec:model} formulates the structural topic model on the
simplex and derives the ILR reparametrisation.
Section~\ref{sec:analytical} presents the analytical solution via
spectral profiling, including the post-estimation rotation and
topic--term refinement.
Section~\ref{sec:theory} establishes the theoretical properties:
global optimality, identifiability, consistency, asymptotic
normality, and the linearisation error bound.
Section~\ref{sec:connections} discusses connections to existing
frameworks.
Section~\ref{sec:algorithm} describes the complete algorithmic
pipeline.
Section~\ref{sec:simulation} reports the simulation study.
Section~\ref{sec:application} presents the British Election Study
application.
Section~\ref{sec:conclusions} concludes.