---
title: "Structural Topic Modeling in Aitchison Geometry: Blog Posts on the 2008 US Presidential Election"
author: "sgscatm package"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Structural Topic Modeling in Aitchison Geometry: Blog Posts on the 2008 US Presidential Election}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

# Introduction

`sgscatm` fits a **Structural Topic Model in Aitchison geometry** via a
closed-form spectral solution.  Document–topic proportions live on the
probability simplex $\mathcal{S}^K$ and are linked to document-level
covariates through a Generalized Structured Component Analysis (GSCA)
formulation.  Estimation reduces to a **single eigendecomposition**—no
iterative algorithm, no random initialisation, no local optima.

This vignette walks through a complete analysis of `poliblog5k`, a corpus of
5 000 US political blog posts collected during the 2008 presidential
campaign and distributed with the `stm` package
(Roberts, Stewart & Tingley, 2019).

## The ILR-Spectral-GSCA model in brief

For $M$ documents, $N$ terms, $K$ topics and $P$ covariates:

| Symbol | Dimension | Meaning |
|--------|-----------|---------|
| $\mathbf{W}$ | $M \times N$ | Document–term matrix |
| $\mathbf{C}$ | $M \times P$ | Covariate matrix (column-centred) |
| $\mathbf{\Pi}$ | $M \times K$ | **Topic proportions** (rows $\in \mathcal{S}^K$) |
| $\boldsymbol{\Phi}$ | $K \times N$ | Topic–term loadings |
| $\mathbf{B}_z$ | $P \times (K{-}1)$ | Covariate path coefficients in ILR space |

The model solves

$$
\min_{\mathbf{Z},\boldsymbol{\Psi},\mathbf{B}_z,\boldsymbol{\mu}}
\left\|
  \widetilde{\mathbf{W}} - \tfrac{1}{K}\,\mathbf{Z}\boldsymbol{\Psi}
\right\|_F^2
+
\lambda\,\left\|\mathbf{Z} - \mathbf{C}\mathbf{B}_z\right\|_F^2,
\qquad
\text{s.t.}\; \mathbf{Z}^\top\mathbf{Z} = \mathbf{I}_{K-1},
$$

where $\mathbf{Z}$ are the ILR topic scores and $\widetilde{\mathbf{W}}$
is the row-centred document–term matrix.  After profiling out
$\boldsymbol{\Psi}$, $\mathbf{B}_z$ and $\boldsymbol{\mu}$, the problem
reduces to the top-$(K{-}1)$ eigendecomposition of the
$M \times M$ matrix

$$
\mathbf{S}_z =
\underbrace{\widetilde{\mathbf{W}}\widetilde{\mathbf{W}}^\top}_{\text{centred doc similarity (LSA)}}
+
\lambda\,\underbrace{\mathbf{C}(\mathbf{C}^\top\mathbf{C})^{-1}\mathbf{C}^\top}_{\text{covariate projector } \mathbf{P}_C}.
$$

$\lambda = 0$ recovers centred LSA; $\lambda \to \infty$ collapses topics
onto the covariate space.

---

# Setup

```{r pkgs, message=FALSE, warning=FALSE}
# Load sgscatm functions (replace with library(sgscatm) once installed)
pkg_dir <- system.file("R", package = "sgscatm")
if (nchar(pkg_dir) == 0L) {
  # running from source
  for (f in list.files("../R", pattern = "[.]R$", full.names = TRUE)) source(f)
} else {
  library(sgscatm)
}

library(stm)   # for data and comparison
set.seed(2024)
```

---

# Data

## The `poliblog5k` corpus

```{r load-data}
data(poliblog5k)

cat("Total documents :", length(poliblog5k.docs), "\n")
cat("Vocabulary size :", length(poliblog5k.voc),  "\n")
cat("Metadata columns:", names(poliblog5k.meta),  "\n")
```

The corpus covers the period from the early 2008 primaries to the
November election.  Each document has two covariates:

- **`rating`** — political leaning of the blog (*Conservative* / *Liberal*).
- **`day`** — day of the campaign (integer, roughly 1–365).

```{r meta-summary}
table(poliblog5k.meta$rating)
```

## Sampling and vocabulary pruning

We draw a random sample of **1 000 documents** and use
`prepDocuments()` to remove terms appearing in fewer than 2 documents
(required by `stm`; we re-use the same pruned vocabulary for `sgscatm`
to ensure a fair comparison).

```{r sample}
N_sample <- 1000L
idx      <- sample.int(5000L, N_sample)

out   <- prepDocuments(
  poliblog5k.docs[idx],
  poliblog5k.voc,
  poliblog5k.meta[idx, ],
  lower.thresh = 2L,
  verbose      = FALSE
)

docs  <- out$documents
vocab <- out$vocab
meta  <- out$meta
M     <- length(docs)
V     <- length(vocab)

cat(sprintf("After pruning: %d documents, %d terms\n", M, V))
```

## Building the document–term matrix for `sgscatm`

`sgscatm` expects a dense numeric matrix $\mathbf{W} \in \mathbb{R}_{\ge 0}^{M \times N}$.
We convert `stm`'s sparse list format:

```{r dtm}
stm_to_dtm <- function(docs, V) {
  W <- matrix(0.0, length(docs), V)
  for (i in seq_along(docs)) {
    d <- docs[[i]]
    W[i, d[1L, ]] <- d[2L, ]
  }
  W
}

W <- stm_to_dtm(docs, V)
cat(sprintf("DTM: %d x %d,  sparsity %.1f%%\n",
            nrow(W), ncol(W), 100 * mean(W == 0)))
```

## Covariate matrix

We use two covariates: political leaning (0/1) and campaign day (centred).

```{r covariates}
C <- scale(
  cbind(rating = as.integer(meta$rating == "Liberal"),
        day    = meta$day),
  center = TRUE, scale = FALSE
)
colnames(C) <- c("Liberal", "day")
head(round(C, 2))
```

---

# Fitting the model

`sgscatm()` requires only $\mathbf{W}$, $\mathbf{C}$, the number of
topics $K$, and the regularisation weight $\lambda$.  The default
`scale_W = TRUE` row-normalises $\mathbf{W}$ to term frequencies before
fitting.

```{r fit, cache=TRUE}
K      <- 7L
lambda <- 3.0

t_start <- proc.time()
fit <- sgscatm(W, C, K = K, lambda = lambda, scale_W = TRUE)
t_fit   <- proc.time() - t_start

cat(sprintf("Fit completed in %.2f seconds (elapsed)\n", t_fit["elapsed"]))
```

## Eigenvalue spectrum

The top $K - 1 = 6$ eigenvalues of $\mathbf{S}_z$ quantify how much
variance in the covariate-augmented document similarity is captured by
each ILR direction.

```{r eigenvalues}
ev_df <- data.frame(
  Direction  = paste0("ILR", seq_len(K - 1L)),
  Eigenvalue = round(fit$eigenvalues, 3),
  Cumulative = round(cumsum(fit$eigenvalues) / sum(fit$eigenvalues_all), 3)
)
knitr::kable(ev_df, align = "lrr",
             caption = "Top eigenvalues of $S_z$")
```

The first two eigenvalues dominate and correspond to the covariate
projector $\lambda \mathbf{P}_C$ (with $\lambda = 3$, this term
contributes up to $3 \times \text{rank}(\mathbf{C}) = 6$ to the trace).
The remaining directions capture purely textual variance (LSA signal).

---

# Topic–term structure

## Top-10 keywords per topic

```{r top-terms}
tt <- top_terms(fit, n = 10L, vocab = vocab)

for (k in seq_len(K)) {
  cat(sprintf("Topic %d : %s\n", k, paste(tt[k, ], collapse = "  |  ")))
}
```

The topics capture coherent themes from the 2008 campaign:

| Topic | Interpretation |
|-------|----------------|
| 1 | War in Iraq / Bush administration |
| 2 | General political commentary |
| 3 | Sarah Palin / financial crisis |
| 4 | Obama–McCain general election |
| 5 | Obama–Clinton Democratic primary |
| 6 | Electoral mechanics (voting, polls, parties) |
| 7 | Republican foreign policy / Iran |

## ILR topic–term deviation matrix

The matrix $\hat{\boldsymbol{\Psi}} \in \mathbb{R}^{(K-1) \times N}$
captures topic-specific word deviations in ILR coordinates.  Each row is
the ILR projection of one direction; positive entries indicate words
*over-represented* relative to the corpus baseline in that direction.

```{r psi-heatmap, fig.width=7, fig.height=3.5}
# Show the 20 most variable terms across ILR directions
term_var <- apply(fit$Psi, 2L, var)
top_var  <- order(term_var, decreasing = TRUE)[1:20]

image(
  t(fit$Psi[, top_var]),
  xaxt = "n", yaxt = "n",
  col  = hcl.colors(64, "Blue-Red 3"),
  main = expression(paste("ILR deviation matrix ", hat(Psi),
                          " (top-20 variable terms)")),
  xlab = "Term", ylab = "ILR direction"
)
axis(1, at = seq(0, 1, length.out = 20),
     labels = vocab[top_var], las = 2, cex.axis = 0.65)
axis(2, at = seq(0, 1, length.out = K - 1L),
     labels = paste0("ILR", seq_len(K - 1L)), las = 1, cex.axis = 0.8)
```

---

# Document–topic proportions

## Distribution across the corpus

`fit$Pi` is an $M \times K$ matrix; each row sums to 1 by construction
(the closure map $\text{softmax}(\mathbf{V}\mathbf{z}_i^*)$ is the
exact inverse of the ILR transformation, not an approximation).

```{r pi-check}
cat("Row-sum range:", range(round(rowSums(fit$Pi), 10)), "\n")
cat("Min proportion:", min(fit$Pi), "\n")
```

```{r pi-boxplot, fig.height=3.5}
boxplot(
  fit$Pi,
  names  = paste0("T", seq_len(K)),
  col    = hcl.colors(K, "Set 2", alpha = 0.7),
  border = hcl.colors(K, "Set 2"),
  ylab   = "Topic proportion",
  main   = "Distribution of document–topic proportions",
  las    = 1
)
abline(h = 1 / K, lty = 2, col = "grey50")
legend("topright", "1/K centroid", lty = 2, col = "grey50", bty = "n",
       cex = 0.8)
```

The median of each topic is close to $1/K \approx 0.14$ — this is
expected: the ILR centroid $\mathbf{z} = \mathbf{0}$ maps exactly to
the uniform composition, so the *average* topic proportion is $1/K$
by construction.  The spread reflects how strongly individual documents
deviate from the centroid.

---

# Covariate effects

## ILR path coefficients $\hat{\mathbf{B}}_z$

$\hat{\mathbf{B}}_z \in \mathbb{R}^{P \times (K-1)}$ gives the OLS
regression of the ILR topic scores on the covariates.  Column $k$
describes how covariate $p$ shifts the topic composition along the $k$-th
ILR direction.

```{r bz}
Bz_df <- as.data.frame(round(fit$Bz, 4))
rownames(Bz_df) <- c("Liberal (vs Conservative)", "day (centred)")
colnames(Bz_df) <- paste0("ILR", seq_len(K - 1L))
knitr::kable(Bz_df,
             caption = "Path coefficients $\\hat{\\mathbf{B}}_z$")
```

The `Liberal` coefficient on ILR1 (−0.063) is the largest signal:
Liberal blogs score systematically lower on the first ILR direction.
Since ILR1 contrasts the first topic against the others, this
reflects a shift in topic composition between political leanings.

## Mean topic prevalence by leaning

```{r rating-effect, fig.height=3.5}
prev_cons <- colMeans(fit$Pi[meta$rating == "Conservative", ])
prev_lib  <- colMeans(fit$Pi[meta$rating == "Liberal",      ])

x   <- seq_len(K)
ylim <- range(c(prev_cons, prev_lib)) + c(-0.005, 0.005)

plot(x - 0.12, prev_cons, pch = 19, col = "#E64B35",
     xlim = c(0.5, K + 0.5), ylim = ylim,
     xlab = "Topic", ylab = "Mean prevalence",
     main = "Mean topic prevalence by political leaning",
     xaxt = "n")
points(x + 0.12, prev_lib, pch = 17, col = "#4DBBD5")
axis(1, at = x, labels = paste0("T", x))
legend("topright",
       legend = c("Conservative", "Liberal"),
       pch    = c(19, 17),
       col    = c("#E64B35", "#4DBBD5"),
       bty    = "n")
abline(h = 1/K, lty = 2, col = "grey70")
segments(x - 0.12, prev_cons, x + 0.12, prev_lib,
         col = "grey60", lty = 3)
```

```{r delta-table}
delta_df <- data.frame(
  Topic        = paste0("T", seq_len(K)),
  Conservative = round(prev_cons, 4),
  Liberal      = round(prev_lib,  4),
  Delta        = round(prev_lib - prev_cons, 4),
  Leaning      = ifelse(prev_lib > prev_cons, "Liberal ↑", "Conservative ↑")
)
knitr::kable(delta_df, align = "lrrrr",
             caption = "Mean prevalence difference (Liberal − Conservative)")
```

## Prevalence over time

```{r day-effect, fig.height=4}
# Bin campaign days into quartiles for visualisation
day_q <- cut(meta$day, breaks = quantile(meta$day, c(0,.25,.5,.75,1)),
             include.lowest = TRUE,
             labels = c("Q1 (early)", "Q2", "Q3", "Q4 (late)"))

prev_time <- t(sapply(levels(day_q), function(q)
  colMeans(fit$Pi[day_q == q, ])))

matplot(prev_time, type = "b", pch = seq_len(K),
        col = hcl.colors(K, "Set 2"),
        xaxt = "n",
        xlab = "Campaign quarter", ylab = "Mean prevalence",
        main = "Topic prevalence across campaign timeline",
        lty  = 1)
axis(1, at = seq_len(4), labels = levels(day_q))
abline(h = 1/K, lty = 2, col = "grey70")
legend("topright", paste0("T", seq_len(K)),
       pch = seq_len(K), col = hcl.colors(K, "Set 2"),
       bty = "n", ncol = 2, cex = 0.8)
```

---

# Prediction on new documents

`predict.sgscatm()` projects new documents onto the fitted topic space
via the ILR loadings $\hat{\boldsymbol{\Psi}}$ without refitting.

```{r predict}
# Hold out 50 documents for prediction
idx_new  <- sample.int(M, 50L)
W_new    <- W[idx_new, ]
Pi_new   <- predict(fit, W_new, scale_W = TRUE)

cat("Predicted Pi dimensions:", dim(Pi_new), "\n")
cat("Row-sum range:", range(round(rowSums(Pi_new), 10)), "\n")

# Top predicted topic per document
top_topic <- apply(Pi_new, 1L, which.max)
cat("Topic distribution in new docs:\n")
print(table(paste0("T", top_topic)))
```

---

# Bootstrap standard errors for path coefficients

`ilr_se()` estimates standard errors for $\hat{\mathbf{B}}_z$ via
nonparametric bootstrap resampling of documents.

```{r se, cache=TRUE}
se_res <- ilr_se(fit, W, C, B = 100L, seed = 42L, verbose = FALSE)

se_df <- as.data.frame(round(se_res$se, 5))
rownames(se_df) <- rownames(Bz_df)
colnames(se_df) <- paste0("ILR", seq_len(K - 1L))
knitr::kable(se_df,
             caption = "Bootstrap SE for $\\hat{\\mathbf{B}}_z$ (B = 100)")
```

```{r ci-plot, fig.height=3.5}
# Plot Bz +/- 1.96*SE for the 'Liberal' covariate (row 1)
bz_liberal <- fit$Bz[1L, ]
se_liberal  <- se_res$se[1L, ]

ci_lo <- bz_liberal - 1.96 * se_liberal
ci_hi <- bz_liberal + 1.96 * se_liberal
dirs  <- seq_len(K - 1L)

plot(dirs, bz_liberal, pch = 19, col = "#3C5488",
     ylim = range(c(ci_lo, ci_hi)) + c(-0.01, 0.01),
     xaxt = "n",
     xlab = "ILR direction", ylab = "Coefficient",
     main = expression(paste(hat(B)[z],
       ": 'Liberal' covariate with 95% bootstrap CI")))
arrows(dirs, ci_lo, dirs, ci_hi,
       length = 0.05, angle = 90, code = 3, col = "#3C5488")
abline(h = 0, lty = 2, col = "grey50")
axis(1, at = dirs, labels = paste0("ILR", dirs))
```

---

# Summary

```{r summary}
print(fit)
```

```{r summary-table, echo=FALSE}
summ_df <- data.frame(
  Parameter        = c("Documents (M)", "Terms (N)", "Topics (K)",
                       "Covariates (P)", "Lambda (λ)", "Fit time (s)"),
  Value            = c(M, V, K, ncol(C), lambda,
                       round(t_fit["elapsed"], 2))
)
knitr::kable(summ_df, align = "lr", caption = "Model summary")
```

## Key properties

- **Non-iterative**: the solution is the top-$(K{-}1)$ eigenspace of an
  $(r + P) \times (r + P)$ matrix ($r \le \min(M, N, 100)$) — no random
  initialisation, no convergence issues, fully deterministic.
- **Globally optimal**: for the profiled, linearised ILR-Spectral-GSCA
  objective (Ky Fan theorem).
- **Simplex-consistent**: topic proportions satisfy
  $\hat{\pi}_{ik} \ge 0$ and $\sum_k \hat{\pi}_{ik} = 1$ by
  construction — not through post-hoc clipping.
- **Compositional**: the structural penalty
  $\lambda \|\mathbf{Z} - \mathbf{C}\mathbf{B}_z\|_F^2$ equals the
  sum of squared **Aitchison distances** between observed and
  covariate-predicted topic compositions (isometry of ILR).

## Comparison with STM (simulation evidence)

The simulation study in `scripts/simulation/run_simulation.R` (Block 4)
benchmarks `sgscatm` against STM across three scenarios and up to
M = 20 000 documents. Key empirical findings relevant to this vignette:

- **Speed**: `sgscatm` is 10–60× faster than STM at all corpus sizes tested
  (e.g. 0.1 s vs 6 s at M = 2 000, K = 5).
- **Accuracy under weak signal**: when the covariate effect is small,
  `sgscatm` recovers $\mathbf{B}_z$ more accurately than STM's variational
  EM (MSE 0.001 vs 0.009 at M = 500).
- **Best of both worlds**: `stm+warm` — STM initialised from the `sgscatm`
  exact solution — achieves the lowest overall MSE(B̂z) in high- and
  low-separation settings, at a fraction of the iterations of vanilla STM
  (10 vs 22 iterations under low topic separation).
- **Topic-word quality**: a single `refine_phi()` step after `sgscatm`
  reduces MSE($\hat{\boldsymbol{\Phi}}$) by ~100× at negligible extra cost,
  closing most of the gap with STM's EM-refined topic-word distributions.

To reproduce: `Rscript scripts/simulation/run_simulation.R --block 4 --quick`.

---

# Session info

```{r session}
sessionInfo()
```
